---
title: "Using Restricted Cubic Splines in Structural Equation Models"
author: "Steven M. Brunwasser, Ph.D."
date: "`r Sys.Date()`"
output:
  html_document:
    theme: sandstone
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(echo = TRUE,
                      warning = F,
                      message = F 
                      )
```


## Workspace Prep

```{r}
require( data.table )
require( ggplot2 )
require( mgcv )
require( rms )
require( Hmisc )
require( lavaan )
require( ERP )
```


## Demonstration in OLS regression
Prior to demonstrating the use of restricted cubic splines (RCSs) in structural equation models (SEM), we will use RCSs to model a nonlinear association in an ordinary least squares (OLS) model using the [rms](). There are several helpful overviews for this package, including [this one](https://www.nicholas-ollberding.com/post/an-introduction-to-the-harrell-verse-predictive-modeling-using-the-hmisc-and-rms-packages/) and [this one](https://www.r-bloggers.com/2016/07/introduction-to-the-rms-package/).  

First, let's simulate data with a nonlinear association that does not follow a simple polynomial function. Goldfield and Wujciak-Jens provide [helpful code](https://www.rdatagen.net/post/2022-08-09-simulating-data-from-a-non-linear-function-by-specifying-some-points-on-the-curve/) that to accomplish this. Please see their code and explanations for a detailed explanation of this approach.
```{r}
x <- c(20, 30, 40, 50, 60, 70, 80)
y <- c(15, 40, 25, 22, 35, 24, 15)

dd <- data.table(x = x, y = y)

dc <- copy(dd)

dc[, xend := (shift(x, type = "lead"))]
dc[, yend := (shift(y, type = "lead"))]
dc <- dc[ -.N]
dc[, id := .I]

dc


interp_line <- function(x, y, xend, yend) {
  slope <- (yend - y)/(xend - x)
  b <- y - slope*x
  newx <- seq(x, xend, length = 100)
  newy <- newx*slope + b
  
  data.table::data.table(x = newx, y = newy)
}

dx <- dc[, interp_line(x, y, xend, yend), keyby = id]
dx

gam.fit <- gam(y ~ s(x, k = 7), data = dx)


dx[, ypred := predict(gam.fit)]


set.seed(123)
mdata <- data.table(x = runif(10000, 20, 85))
mdata$y <- predict(gam.fit, newdata = mdata) + rnorm(nrow(mdata),0, 2)

ggplot(data = mdata, aes(x = x, y = y)) +
  geom_point(size = 1)  +
  scale_x_continuous( breaks = seq( 20, 80, 5 ) )

```


```{r}

dd <- datadist( mdata )
options( datadist = 'dd' ) 

ols.rcs5 <- ols( y ~ rcs( x, 6 ),
            data = mdata,
            x = T,
            y = T )
print( ols.rcs5 )


```

```{r}
anova( ols.rcs5 )
```



```{r}
ols.pol5 <- ols( y ~ pol( x, 5 ),
            data = mdata,
            x = T,
            y = T )
print( ols.pol5 )
anova( ols.pol5, tol=1e-15 )

```


```{r}
ols.rcs5.pred <- Predict( ols.rcs5, x ) 
ols.pol5.pred <- Predict( ols.pol5, x ) 

ols5.pred <- rbind( ols.rcs5.pred, ols.pol5.pred )
ols5.pred$model <- rep( c( '6-Knot RC Spline','Quintic Polynomial' ), each = 200 )

cbbPalette <- c("#E69F00","#0072B2" , "#56B4E9", "#009E73", "#F0E442", "#D55E00", "#CC79A7")

ols.pred5.plot <- ggplot( ) +
  geom_point( data=mdata, aes( x=x, y=y ), color = 'darkgray', shape = 1 ) +
  geom_line( data=ols5.pred, aes( x=x, y=yhat, group = model, colour=model, linetype = model ), size = 2 ) +
  scale_color_manual( values=cbbPalette ) +
  geom_vline( xintercept = rcspline.eval( mdata$x, nk = 6, knots.only = T ), linetype = 3 ) +
  scale_x_continuous( breaks = rcspline.eval( mdata$x, nk = 6, knots.only = T ),
                      labels = c('23.15\nKnot 1','35.06\nKnot 2','46.74\nKnot 3','58.04\nKnot4','69.63\nKnot5','81.61\nKnot 6')) +
  labs( colour='Model Type',
        linetype ='Model Type') +
  theme( axis.text.x =  element_text( size = 12, face = 'bold'),
         legend.title = element_text( size = 14, face = 'bold'),
         legend.text = element_text( size = 12 ),
         axis.title = element_text( size = 14, face = 'bold'),
         legend.position = 'bottom' )
ols.pred5.plot

```
<br>

## GLS Model with ERP Data

```{r}
data(impulsivity)
dim(impulsivity)

impulsivity1 <- subset( impulsivity, !duplicated( impulsivity$Subject ) )

imp <- reshape( impulsivity1,
                direction = 'long',
                timevar = 'time',
                varying = list( out = colnames( impulsivity )[5:505] ),
                idvar = 'Subject',
                times = seq( 0, 1000, 2 )
                )

```


```{r}
ddimp <- datadist( imp )
options( datadist = ddimp )



# imp.rcs7 <- Gls( T_0 ~ rcs( time, 8 ),
#                  correlation = corAR1( form = ~time | Subject  ),
#                  data = imp,
#                  x = T ) 
# save( imp.rcs7, file='imp.rcs7.RData' )
load( 'imp.rcs7.RData' )

# imp.rcs6 <- Gls( T_0 ~ rcs( time, 7 ),
#                  correlation = corAR1( form = ~time | Subject  ),
#                  data = imp,
#                  x = T )
# save( imp.rcs6, file='imp.rcs6.RData' )
load( 'imp.rcs6.RData' )

# imp.rcs5 <- Gls( T_0 ~ rcs( time, 6 ),
#                 correlation = corAR1( form = ~time | Subject  ),
#                  data = imp,
#                  x = T )
# 
# save( imp.rcs5, file='imp.rcs5.RData' )
load( 'imp.rcs5.RData' )

imp.rcs7.aic <- AIC( imp.rcs7 )
imp.rcs6.aic <- AIC( imp.rcs6 )
imp.rcs5.aic <- AIC( imp.rcs5 )

data.frame( AIC = c(imp.rcs7.aic, imp.rcs6.aic, imp.rcs5.aic ),
            DF = c( 7, 6, 5 ) )
```


```{r}
print( imp.rcs6 )
anova( imp.rcs6 )
```

```{r}
# imp.pol6 <- Gls( T_0 ~ pol( time, 6 ),
#                  correlation = corAR1( form = ~time | Subject  ),
#                  data = imp,
#                  x = T )
# save( imp.pol6, file='imp.pol6.RData' )
load( 'imp.pol6.RData' )
```


```{r}
imp.rcs6.pred <- data.frame( Predict( imp.rcs6, time ) )
imp.rcs6.pred$model <- '7-Knot RC Spline'
imp.pol6.pred <- data.frame( Predict( imp.pol6, time ) )
imp.pol6.pred$model <- 'Sextic Polynomial'

imp6.pred <- rbind( imp.rcs6.pred, imp.pol6.pred )
imp6.pred$model <- factor( imp6.pred$model )

impplot <- ggplot( ) +
  geom_line( data=imp, aes( x=time, y=T_0, group = Subject  ), colour='darkgray' ) +
  geom_line( data=imp6.pred, aes( time, yhat, group = model, colour = model, linetype=model ), size=1 ) +
  geom_ribbon( data=imp6.pred, aes( x=time, ymin=lower, ymax=upper, fill=model ), alpha = .3 ) +
  geom_vline( xintercept = rcspline.eval( imp$time, nk = 7, knots.only = T ), linetype = 3 ) +
  scale_color_manual( values=cbbPalette) +
  scale_fill_manual( values=cbbPalette) +
  scale_x_continuous( breaks = rcspline.eval( imp$time, nk = 7, knots.only = T ),
                      labels = c('24\nKnot 1','182\nKnot 2','342\nKnot 3','500\nKnot 4','658\nKnot 5','818\nKnot 6','976\nKnot 7')) +
  labs( y='ERP Amplitude', x='Time (ms)', colour = 'Model Type', fill = 'Model Type', linetype = 'Model Type') +
  theme( legend.position = 'bottom' )
impplot
```

## Latent Growth Curve Model: Perceived Stress Trajectories

```{r}
load( 'cts.splines.Rdata' )
html( describe( cts.splines ) )



```

Create a wide version of the CTS data for latent growth modeling.
```{r}
cts.wide <- reshape( cts.splines,
                     direction = 'wide',
                     v.names = 'psstot',
                     timevar = 'weeks',
                     idvar = 'id',
                     sep = ''
                     )
```


We do not want to make the assumption that the time-invariant dysfunctional attitudes variable (*dasct*) affects stress trajectories in a linear manner; rather, we would like to at least allow for possible non-linearity. We'll model the effect of *dasct* with a 3-knot restricted cubic spline allowing the knots to be placed at their default values (outer knots placed at the 0.10 and 0.90 quantiles of *dasct*). There is no equivalent to the **rcs()** function provided in the *rms* package in the *lavaan* package. Consequently, we need to create $k-1$ time variables (where $k$ is the number of knots) to include in the *cts.wide* data frame in order to estimate the nonlinear effect. We use the **rcspline.eval()** function (rms package) to create new nonlinear time variable (*dasct.nonlinear*) needed to estimate a 3-knot restricted cubic spline effect of dysfunctional attitudes on stress trajectories, then creating a new data frame (*cts.wide1*) that merges this new nonlinear variable with the *cts.wide* data frame. We can then estimate the 3-knot restricted cubic spline effect by including both the *dasct* and the *dasct.nonlinear* variables as predictors in the model, representing the linear and nonlinear effects, respectively.
```{r}
das.spline <- data.frame( rcspline.eval( cts.wide$dasct, nk = 3 ) )
colnames( das.spline ) <- c( 'dasct.nonlinear' )

cts.wide1 <- cbind( cts.wide, das.spline )
head( cts.wide1 )

```


We also want to use a 3-knot restricted cubic spline to estimate a nonlinear stress trajectory. Unlike in standard regression models where time effects are captured by observed variables included in the dataset, time effects are captured by latent (unobserved) intercept & slope variables ("growth factors") in latent growth curve (LGC) models. The factor loadings are (typically) fixed at values selected to model a specific trajectory shape. To model a nonlinear trajectory using a restricted cubic spline, we need to have $k-1$ latent slope variables to capture the time effect; so we need 2 latent slope factors to estimate a 3-knot restricted cubic spline. Note, this is the same number of latent slopes needed to estimate a quadratic growth curve model. The question is: how do we constrain the factor loadings in order to estimate the appropriate restricted cubic spline trajectory. Again, the **rcspline.eval()** function can help us so that we don't have to figure it manually.  
```{r}
knots <- data.frame( rcspline.eval( cts.splines$weeks, nk = 3, knots = c( 4, 8, 14 ), inclx = T ) )
knots1 <- knots[ !duplicated( knots$x ), ] 
colnames( knots1 ) <- c('Linear Slope','Nonlinear Slope' )
knots1

```


```{r}
stress.lgc1 <- '
## Define the latent growth factors -- intercept, linear slope, and nonlinear slope
i =~ 1*psstot0 + 1*psstot2 + 1*psstot4 + 1*psstot6 + 1*psstot8 + 1*psstot10 + 1*psstot12 + 1*psstot13
lin =~ 0*psstot0 + 2*psstot2 + 4*psstot4 + 6*psstot6 + 8*psstot8 + 10*psstot10 + 12*psstot12 + 13*psstot13
nonlin =~ 0*psstot0 + 0*psstot2 + 0*psstot4 + 0.08*psstot6 + 0.64*psstot8 + 2.026667*psstot10 + 4.053333*psstot12 + 5.206667*psstot13
## Estimate first-order autoregressive effect -- hold effect constant over time by giving each the same label -- phi
# psstot13 ~ phi*psstot12
# psstot12 ~ phi*psstot10
# psstot10 ~ phi*psstot8
# psstot8 ~ phi*psstot6
# psstot6 ~ phi*psstot4
# psstot4 ~ phi*psstot2
# psstot2 ~ phi*psstot0
## Constrain variances of latent growth models to = 0 -- i.e., no random effects
i ~~ i
lin ~~ lin
nonlin ~~ nonlin
i ~~ lin
i ~~ nonlin
lin ~~ nonlin
## Constrain error variances to be constant over time
psstot0 ~~ e*psstot0
psstot2 ~~ e*psstot2
psstot4 ~~ e*psstot4
psstot6 ~~ e*psstot6
psstot8 ~~ e*psstot8
psstot10 ~~ e*psstot10
psstot12 ~~ e*psstot12
psstot13 ~~ e*psstot13
## Estimate effects of DAS on stress trajectories
i  ~ dasct + dasct.nonlinear
' 

stress.lgc1.fit <- growth( stress.lgc1,
                          data = cts.wide1,
                          estimator = 'ml' )
summary( stress.lgc1.fit, fit.measures = T )
```




```{r}
stress.lgc2 <- '
## Define the latent growth factors -- intercept, linear slope, and nonlinear slope
i =~ 1*psstot0 + 1*psstot2 + 1*psstot4 + 1*psstot6 + 1*psstot8 + 1*psstot10 + 1*psstot12 + 1*psstot13
lin =~ 0*psstot0 + 2*psstot2 + 4*psstot4 + 6*psstot6 + 8*psstot8 + 10*psstot10 + 12*psstot12 + 13*psstot13
## Estimate first-order autoregressive effect -- hold effect constant over time by giving each the same label -- phi
# psstot13 ~ phi*psstot12
# psstot12 ~ phi*psstot10
# psstot10 ~ phi*psstot8
# psstot8 ~ phi*psstot6
# psstot6 ~ phi*psstot4
# psstot4 ~ phi*psstot2
# psstot2 ~ phi*psstot0
## Constrain variances of latent growth models to = 0 -- i.e., no random effects
i ~~ i
lin ~~ lin
i ~~ lin
## Constrain error variances to be constant over time
psstot0 ~~ e*psstot0
psstot2 ~~ e*psstot2
psstot4 ~~ e*psstot4
psstot6 ~~ e*psstot6
psstot8 ~~ e*psstot8
psstot10 ~~ e*psstot10
psstot12 ~~ e*psstot12
psstot13 ~~ e*psstot13
## Estimate effects of DAS on stress trajectories
i  ~ dasct + dasct.nonlinear
' 

stress.lgc2.fit <- growth( stress.lgc2,
                          data = cts.wide1,
                          estimator = 'ml' )
summary( stress.lgc2.fit, fit.measures = T )
```




```{r}
anova( stress.lgc1.fit, stress.lgc2.fit )
```








```{r}

dd.cts <- datadist( cts.splines )
options( datadist = 'dd.cts')
gls.stress1 <- Gls( psstot ~ rcs( weeks, 3),
                    correlation = corAR1( form =~ weeks | id ),
                    data = cts.splines,
                    x = T )
summary( gls.stress1 )
print( gls.stress1 )
```


```{r}
lme.stress1 <- lme( psstot ~ rcs( weeks, 3) + rcs( dasct, 3 ),
                    random =~ rcs( weeks, 3 ) | id,
                    data = cts.splines,
                    na.action = 'na.omit',
                    method = 'ML')
summary( lme.stress1 )

```





